---
title: "DAB501_Project2_AyeniEzekiel_778486"
output:
  html_document:
    df_print: paged
---

This project displays Univariate and Bivariate Analysis of both numerical and categorical variables.It highlights measure of 
spread and central tendency as well as transformations with respect to different types of distributions. 


INFORMATION ABOUT THE DATASET:

Affairs dataset describes the infidelity and extramarital affairs. It is a cross-section data from a survey conducted by psychology
today in 1969. It is a data frame containing 601 observations on 10 variables. Affairs is an AER package from datasets in R packages.


References:

https://www.statisticshowto.com/multimodal-distribution/

https://www.datasciencemadesimple.com/rounding-off-the-column-in-r-2/



Install Packages:

```{r}
install.packages("tidyverse",repos="http://cran.us.r-project.org")
```


```{r}
library(tidyverse)
```


install AER packages

```{r}
install.packages("AER",repos="http://cran.us.r-project.org")
```


```{r}
library(AER)
```


import dataset: Affairs

```{r}
Affairs <- read.csv("C:/Users/user/OneDrive/Desktop/DAB501/Data/Affairs.csv")
```


Cleaning of Data: 
The first column in the Affairs dataset has no variable name, hence it will be deleted. Also, the yearsmarried variable is in 3 decimal places, it needs to be round off to 1 decimal place because it represent number of years.

```{r}
Affairs_clean<-Affairs %>% select(affairs, gender, age, yearsmarried, children, religiousness, education, occupation, rating)
```


```{r}
Affairs_clean<-Affairs %>% mutate_at(vars(yearsmarried), funs(round(., 1)))
```
 
 
 View clean data
```{r}
View(Affairs_clean)
```
 



1.1  UNIVARIATE ANALLYSIS (NUMERIC VARIABLE): 
```{r}
ggplot(Affairs_clean, aes(x=age)) +geom_histogram(binwidth =4, colour ='white', fill='skyblue')+labs(title="Univariate Analysis with Numeric Variable",x="Age")
```



1.2 OUTLIERS: 
From the histogam plot using 4 as binwidth, it shows suspected outliers for 46yrs to 58yrs of age and for 17.5yrs to 22yrs of age. This was identified by using filter funtion for ages below 20yr and above 45yrs of age. While the outliers from 46yrs and above are reasonable and form part of the distribution because married individuals of such age can still have extra marital affairs, It is therefore surprising to see individual of age 17.5yrs who had been married for the past 10 years (meaning they were married since the age 7.5yrs) having an affair. This outlier should be deleted or neglected as it may have arrise as an error during computation.
```{r}
Outliers<-Affairs_clean%>%filter(age<20)
```

Outliers
```{r}
View(Outliers)
```


1.3 SHAPE AND SKEWNESS OF DISTRIBUTION: 
The shape of the Numerical distribution is Right Skewed, Unimodal with Unusual observations (outliers)


1.4 aPPLY TRANSFORMATION TO DATA
No further transformation is needed for the data, the data has been clean to format the yearsmarried column. The proportion and the count of variables are within reasonable range and as such no log transformation will be required for the data.



1.5 CHOOSE AND CALCULATE APPROPRIATE MEASSURE OF CENTRAL TENDENCY:
For measure of central tendency,  we will use median for calculation

```{r}
median(Affairs_clean$age)
```



1.6 REASON FOR MEDIAN AS MEASURE OF CENTRAL TENDENCY: 
This is because the distribution is skewed to the right. which means that mean is greater than median for the distribution. The distribution also has some extreme outliers. An example is shown below.
```{r}
ggplot(Affairs_clean, aes(x=age)) +geom_histogram(binwidth =4, color ='white', fill='skyblue')+labs(title="Visualizing Central Tendency using Median",x="Age") + geom_vline(aes(xintercept = median(age), color='yellow'))
```


1.7 MEASURE OF SPREAD: 
For measure of spread,  we will use Interquartile Range (IQR) for calculation. This is because IQR is more helpful and more robust to skewness and outliers in describing the center sprea

```{r}
IQR(Affairs_clean$age)
```





2.1  UNIVARIATE ANALLYSIS (CATEGORICAL VARIABLE): 
     Create an appropriate plot to visualize the distribution of counts for this variable.
     
```{r}
ggplot(Affairs_clean, aes(x=education)) + geom_bar( fill = 'skyblue', colour='black')+labs(title="Univariate Analysis using Categrical Variable for distribution of counts",x="Level of Education")
```
     
     
     
     
2.2 Create an appropriate plot to visualize the distribution of proportions for this variable.

```{r}
ggplot(Affairs_clean, aes(x=education, y = ..prop.., grooup = 1), stat = 'count') + geom_bar( fill = 'red', colour='yellow')+labs(title="Univariate Analysis using Categrical Variable for distribution of proportions",x="Level of Education")
```



2.3 Discuss any unusual observations for this variable:

The unusual observation here is that the count of individuals having extramarital affairs happens to come from the learned cadre as against the expectation that low level of education will impact sensitization on the dangers of having extra affairs. The distribution depicts that learned people are more engaged in in extra affairs.




2. 4 Discuss if there are too few/too many unique values

The proportion of individuals with basic or elementary education who are engeged in extra affairs are just a little above zero, while individuals with mmiddle or college education tend to be more engaged in extra affairs. The proportion of people with highest level of edcation tends to dwindle when it comes to having extra affairs.




3.1  BIVARIATE ANALYSIS (Numeric Vs Numeric variables)

```{r}
ggplot(Affairs_clean, aes(x=affairs, y=age)) + geom_point(colour='red')+labs(title="Bivariate Analysis Numerical Vs Numerical Variables",x="Extra Affairs", y="Age of Individual")
```


3.2 Describe the form, direction, and strength of the observed relationship. Include both qualitative and quantitative measures, as appropriate. There is no relationship between the scatter plot. This simply means that age is not a factor that correclate or drives extra affairs. Looking at the strenth of the plot, majority of the age group are classified as having more extra affairs as well mininum number of affairs.
```{r}
summary(Affairs_clean$affairs)
```



3.3 Explain what this relationship means in the context of the data.

From the calculation, we can see the minimum value, Q1,Q5 and median to be zero. Only Mean=1.056 and max 12.00 has reasonable values. This is wy there is no strong relationship between the variables. The correlation is dimensionless.


3.4 Describe the variability that you observe in the plot and how that corresponds to the strength you calculated in #2 above.
The tendency for Age to drive extra affairs is not viable as clearly evident in the plot.




4.1 BIVARIATE ANALYSIS (categorical Vs Numeric variables)

```{r}
ggplot(Affairs_clean, aes(x=gender, y=affairs)) + geom_boxplot()+labs(title="Bivariate Analysis categorical Vs Numerical Variables",x="Sex", y="Mumber of Affairs")
```



4.2 Describe the form, direction, and strength of the observed relationship. Include both qualitative and quantitative measures, as appropriate. This analysis is describing how gender drives the number of extra affairs among individuals. there are no significant difference in how the distribution drives extra affairs among both sexes. However, both sexes have significant outliers in the number of extra affairs but the median between both sexes are the same. However, the 75th percentile for extra affairs for male is quite significant. the relationship is week.

4.3. Explain what this relationship means in the context of the data.
Both male and female share the same median i.e 50th percentile in the boxplot. They both have outliers.



4.4 Describe the variability that you observe in the plot and how that corresponds to the strength you calculated in #2 above.
The graph depics that we have lot of male and female with zero engagement in extra affairs. This is why the median (50% percentile) in the box plot tends to zero since median is the measure of value in the middle when arranged in ascending order. However we have some male and female who are engaged in a typical extra affairs with high frequency. This is why there are whiskers and ouliers in the plot.
